\vspace{-1.5cm}\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}

### 2025-05-18 paper/main.tex
next iteration, more concise

\title{\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
\author{Wout De Rijck, Prof. Dr. Ir. Thomas Demeester, \\ Prof. Dr. Ir. Chris Develder, Karel D'Oosterlinck}
\date{} % No date

@@

\begin{abstract}

Large language models (LLMs) require substantial computational and memory resources, limiting their utility in resource-constrained environments. 
ModernBERT is a smaller encoder-only language model that excels at various tasks while being computationally highly efficient. 
In this work, we study how much more ModernBERT can be compressed while retaining accuracy on any given task.
Specifically, we introduce a series of very simple vocabulary pruning techniques that target the embedding layer. We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining on average 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model. Additionally, our results suggest that state-of-the-art encoder-layer pruning can fall short of simple embedding-layer pruning.

\end{abstract}

@@

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements. Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining. Model compression strategies for LLMs can be categorized into four primary methods:

@@

\newpage
\section{Method}
The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters. This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance. The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.

@@

\subsubsection{Frequency-Based}
Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first. This approach operates on the principle that rarely used tokens contribute less to model performance. The technique-specific algorithm steps are:

\begin{enumerate}
    \item Count occurrences of each token in the task-specific dataset
    \item Sort tokens by frequency (most common first)
\end{enumerate}
\subsubsection{Attention-Based}
This method leverages the attention mechanism of transformer models to identify important tokens. The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective. Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.

For example, the important tokens identified in the SST-2 dataset differ markedly between methods. Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
% \begin{figure*}[t]
% \centering
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
% \end{minipage}%
% \hfill
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
% \end{minipage}
% \caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
% \label{fig:wordcloud-comparison}
% \end{figure*}
% The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item First fine-tune the base model on the target task to learn task-specific attention patterns
    \item Process the task dataset through this fine-tuned model, capturing attention matrices from all layers and heads
    \item For each token, aggregate the attention it receives across all contexts where it appears
    \item Normalize attention scores by token frequency to avoid bias toward common tokens
\end{enumerate}
\subsubsection{TF-IDF Based}
Tokens are ranked using Term Frequency-Inverse Document Frequency (TF-IDF) scores, which balance token occurrence frequency with discriminative power across documents. This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.

The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes. Unlike simple frequency counts, TF-IDF prevents common but semantically shallow tokens (articles, conjunctions, etc.) from dominating the pruned vocabulary.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item Tokenize all examples in the task dataset
    \item Calculate term frequency (TF) for each token in each document:
        \begin{center}
        $\text{TF}(t,d) = \frac{\text{count of $t$ in $d$}}{\text{total terms in $d$}}$
        \end{center}
    \item Calculate inverse document frequency (IDF) across the corpus:
        \begin{center}
        $\text{IDF}(t) = \log\left(\frac{\text{total docs}}{\text{docs with $t$}}\right)$
        \end{center}
    \item Compute TF-IDF scores by multiplying TF and IDF values:
        \begin{center}
        $\text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)$
        \end{center}
    \item Apply normalization to the scores (see variants below)
\end{enumerate}
For this approach, three normalization variants were implemented and evaluated:
\begin{itemize}
    \item Non-normalized TF-IDF: Raw scores that heavily prioritize rare but task-specific tokens
    \item L1-normalized TF-IDF: Scores normalized by sum, balancing frequency and discriminative power
    \item L2-normalized TF-IDF: Scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating
\end{itemize}
\subsection{Out-of-Vocabulary Token Handling}
The handling of out-of-vocabulary (OOV) tokens presents an opportunity to preserve additional semantic information during vocabulary pruning. While standard approaches typically map all pruned tokens to a single UNK token, this approach utilizes the semantic properties of the embedding space to maintain some of the original meaning of pruned tokens. This can help mitigate performance degradation by retaining partial semantic value for tokens not included in the reduced vocabulary.
\\ \\
The clustering-based OOV handling process works as follows:
\begin{enumerate}
    \item Extract embeddings for all tokens marked for removal during pruning
    \item Apply K-means clustering to these embeddings, grouping semantically similar tokens together
    \item For each cluster, identify the token closest to the centroid as the representative
    \item Create a mapping from each pruned token to its cluster representative
    \item During inference, when an OOV token is encountered, it is dynamically mapped to its assigned representative rather than the generic UNK token
\end{enumerate}
This approach maintains semantic nuance by ensuring that pruned tokens are replaced by semantically similar alternatives rather than a single catchall token. For example, domain-specific terminology might be mapped to more common synonyms, or rare morphological variants might be mapped to their common forms (e.g., "stabilized" → "stable"). The method is particularly effective for technical vocabularies where semantically related terms form natural clusters in the embedding space.

@@

\section{Conclusion}

% This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact. The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 97.6\% of original model performance while reducing total parameters by 20-23\%.

% The key contributions of this research are:

% \begin{itemize}
%     \item A vocabulary pruning methodology that operates as a pre-fine-tuning optimization step, eliminating the need for post-pruning recovery fine-tuning typically associated with model compression.

%     \item A systematic comparison of token importance estimation techniques, revealing that TF-IDF with OOV handling provides the most robust performance across diverse NLP tasks.

%     \item Empirical evidence that vocabulary pruning outperforms encoder-based pruning at lower pruning ratios (5-20\%), offering a more efficient pathway for moderate model compression.

%     \item Practical deployment benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory with no inference time penalty.
% \end{itemize}

% The results challenge the conventional wisdom that encoder-layer pruning should be prioritized over vocabulary reduction. Instead, the findings demonstrate that a hybrid approach—focusing first on vocabulary pruning for initial efficiency gains and then applying encoder pruning for more aggressive compression—offers the most balanced pathway to efficient transformer models.

% The broader implications of this work extend beyond the specific techniques presented. By eliminating the need for expensive post-pruning fine-tuning, the presented approach democratizes model compression, making it accessible even for organizations with limited computational resources. This could accelerate the adoption of transformer models in resource-constrained environments such as mobile devices, edge computing platforms, and regions with limited technological infrastructure.

% Future work could explore dynamic vocabulary pruning that adapts to changing data distributions, application of these techniques to multilingual models where vocabulary redundancy is even more pronounced, and integration with quantization and knowledge distillation for more comprehensive compression pipelines. As large language models continue to grow in size and capability, efficient adaptation methods like vocabulary pruning will become increasingly critical for bridging the gap between research advances and practical deployment.

\section{Future Work}
TODO (these are just things I thought about)

\begin{itemize}
    \item \textbf{Unified Compression Framework}: Developing an integrated approach that combines vocabulary pruning with encoder-layer compression techniques like LoSparse could provide a more comprehensive model compression solution. Preliminary experiments suggest these approaches are complementary, with vocabulary pruning excelling at lower compression ratios and encoder pruning maintaining better performance at higher ratios.

    \item \textbf{Task Adaptation Networks}: Instead of removing pruned token embeddings entirely, future work could explore low-dimensional adapter networks that efficiently transform a small set of base embeddings to approximate the full vocabulary, potentially offering better performance/size tradeoffs.

    \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.

\end{itemize}
\newpage
\bibliographystyle{plain}
\bibliography{references}

>>>>

\setcounter{secnumdepth}{0}

\title{\vspace{-1.5cm}\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
\author{
Wout De Rijck\\[0.5em]
\small{Supervisors:} \normalsize Prof. Dr. Ir. Thomas Demeester, \normalsize Prof. Dr. Ir. Chris Develder\\
\small{Councellor:} \normalsize Dr. Ir. Karel D'Oosterlinck
}
\date{} % No date

@@

\textbf{Abstract:} Large language models (LLMs) require substantial computational and memory resources, limiting their utility in resource-constrained environments. 
ModernBERT is a smaller encoder-only language model that excels at various tasks while being computationally highly efficient. 
In this work, we study how much more ModernBERT can be compressed while retaining accuracy on any given task.
Specifically, we introduce a series of very simple vocabulary pruning techniques that target the embedding layer. We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining on average 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model. Additionally, our results suggest that state-of-the-art encoder-layer pruning can fall short of simple embedding-layer pruning.

\textbf{Keywords:} vocabulary pruning, ModernBERT, encoder-only model, model compression, embedding-layer, task-specific pruning, TF-IDF

@@

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements. Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining.
Model compression strategies for LLMs can be categorized into four primary methods \cite{wan2023efficient}:

@@

\section{Method}
The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters. This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance. The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.

@@

\subsubsection{Frequency-Based}
Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first. This approach operates on the principle that rarely used tokens contribute less to model performance.

%For example, the important tokens identified in the SST-2 dataset differ markedly between methods. Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
% \begin{figure*}[t]
% \centering
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
% \end{minipage}%
% \hfill
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
% \end{minipage}
% \caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
% \label{fig:wordcloud-comparison}
% \end{figure*}
% The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.

\subsubsection{Attention-Based}
This method leverages the attention mechanism of transformer models to identify important tokens. The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective. Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item First fine-tune the base model on the target task to learn task-specific attention patterns
    \item Process the task dataset through this fine-tuned model, capturing attention matrices from all layers and heads
    \item For each token, aggregate the attention it receives across all contexts where it appears
    \item Normalize attention scores by token frequency to avoid bias toward common tokens
\end{enumerate}
\subsubsection{TF-IDF Based}
Tokens are ranked using Term Frequency-Inverse Document Frequency (TF-IDF) scores, which balance token occurrence frequency with discriminative power across documents. This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.

The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes. Unlike simple frequency counts, TF-IDF prevents common but semantically shallow tokens (articles, conjunctions, etc.) from dominating the pruned vocabulary.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item Tokenize all examples in the task dataset
    \item Calculate term frequency (TF) for each token in each document:
        \begin{center}
        $\text{TF}(t,d) = \frac{\text{count of $t$ in $d$}}{\text{total terms in $d$}}$
        \end{center}
    \item Calculate inverse document frequency (IDF) across the corpus:
        \begin{center}
        $\text{IDF}(t) = \log\left(\frac{\text{total docs}}{\text{docs with $t$}}\right)$
        \end{center}
    \item Compute TF-IDF scores by multiplying TF and IDF values:
        \begin{center}
        $\text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)$
        \end{center}
    \item Apply normalization to the scores (see variants below)
\end{enumerate}
For this approach, three normalization variants were implemented: Non-normalized TF-IDF (raw scores), L1-normalized TF-IDF (scores normalized by sum), and L2-normalized TF-IDF (scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating). Only the L2-normalized variant was evaluated in our experiments.
\subsection{Out-of-Vocabulary Token Handling}
The handling of out-of-vocabulary (OOV) tokens presents an opportunity to preserve additional semantic information during vocabulary pruning. While standard approaches typically map all pruned tokens to a single UNK token, this approach utilizes the semantic properties of the embedding space to maintain some of the original meaning of pruned tokens. This can help mitigate performance degradation by retaining partial semantic value for tokens not included in the reduced vocabulary.
\\ \\
The clustering-based OOV handling process works as follows:
\begin{enumerate}
    \item Extract embeddings for all tokens marked for removal during pruning
    \item Apply K-means clustering to these embeddings, grouping semantically similar tokens together
    \item For each cluster, identify the token closest to the centroid as the representative
    \item Create a mapping from each pruned token to its cluster representative
    \item During inference, when an OOV token is encountered, it is dynamically mapped to its assigned representative rather than the generic UNK token
\end{enumerate}
This approach maintains semantic nuance by ensuring that pruned tokens are replaced by semantically similar alternatives rather than a single catchall token. For example, domain-specific terminology might be mapped to more common synonyms, or rare morphological variants might be mapped to their common forms (e.g., "stabilized" → "stable"). The method is particularly effective for technical vocabularies where semantically related terms form natural clusters in the embedding space.

@@

\section{Conclusion}
In this work, we designed and evaluated a series of vocabulary pruning techniques for the ModernBERT language model to reduce model size while maintaining task performance.

We found that simple vocabulary pruning techniques outperform encoder-layer pruning for moderate compression ratios. Our TF-IDF-based method with OOV handling achieved 97.6\% of the original model's performance across GLUE tasks while reducing total parameters by 20.02\%, compared to LoSparse's 92.9\% retention at the same compression ratio. This resulted in practical benefits of 14.83\% reduction in GPU memory requirements and 20\% less storage space.

These results challenge the assumption that encoder pruning should precede vocabulary reduction and highlight the significant redundancy within embedding layers of pre-trained models, providing an efficient path to model compression for resource-constrained environments.

\section{Future Work}
Future research could explore an integrated approach combining vocabulary pruning with encoder-layer compression techniques like LoSparse, as preliminary experiments suggest these approaches are complementary. We hypothesize that pruning up to a certain threshold using this combined approach will yield better results than using either vocabulary pruning or encoder-layer pruning alone.

Additionally, evaluating vocabulary pruning on information retrieval tasks may reveal different optimal pruning strategies, as retrieval models likely depend on different token importance distributions than classification tasks.

\bibliographystyle{plain}
\bibliography{references}
